---
title: "ST2195 Coursework Part 1"
author: "Richard Hardy"
date: "`26/3/25`"
output: html_document
---

# Introduction

This report implements a Metropolis-Hastings algorithm in R to draw samples from a target distribution.
The analysis replicates the original work done in Python and includes diagnostics such as $\hat{R}$ values across varying proposal scales.

# Setup

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
```

# Part (a): Metropolis-Hastings Sampling

```{r}
# Define target density function
f <- function(x) {
  0.5 * exp(-abs(x))
}

# Log of target density for numerical stability
log_f <- function(x) {
  -abs(x) - log(2)
}

# Metropolis-Hastings algorithm
metropolis_hastings <- function(N = 10000, s = 1, x0 = 0) {
  samples <- numeric(N)
  samples[1] <- x0
  
  for (i in 2:N) {
    x_star <- rnorm(1, mean = samples[i - 1], sd = s)
    log_r <- log_f(x_star) - log_f(samples[i - 1])
    if (log(runif(1)) < log_r) {
      samples[i] <- x_star
    } else {
      samples[i] <- samples[i - 1]
    }
  }
  return(samples)
}

# Run the sampler
set.seed(123)
samples <- metropolis_hastings()

# Sample statistics
sample_mean <- mean(samples)
sample_sd <- sd(samples)
cat("Sample Mean:", sample_mean, "\n")
cat("Sample Standard Deviation:", sample_sd, "\n")
```

```{r, fig.width=8, fig.height=5}
# Plot histogram with target density
x_vals <- seq(-10, 10, length.out = 1000)
f_x <- f(x_vals)
f_df <- data.frame(x = x_vals, y = f_x)

hist_data <- data.frame(samples = samples)

ggplot(hist_data, aes(x = samples)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "deepskyblue", color = "black") +
  geom_line(data = f_df, aes(x = x, y = y), color = "red", size = 1.2) +
  labs(title = "Metropolis-Hastings Sampling (N = 10000, s = 1)", x = "x", y = "Density") +
  theme_minimal() +

  labs(title = "Metropolis-Hastings Sampling (N = 10000, s = 1)",
       x = "x", y = "Density") +
  theme_minimal()
```

# Part (b): R-hat Diagnostic Across Proposal Scale

```{r}
calculate_r_hat <- function(chains) {
  J <- length(chains)
  N <- length(chains[[1]])
  
  M_j <- sapply(chains, mean)
  V_j <- sapply(chains, var)
  
  W <- mean(V_j)
  M <- mean(M_j)
  B <- var(M_j)
  
  R_hat <- sqrt((B + W) / W)
  return(R_hat)
}

# Run chains with varying proposal scale
set.seed(123)
N <- 2000
s_values <- seq(0.001, 1, length.out = 50)
J <- 4
r_hat_values <- numeric(length(s_values))

for (i in seq_along(s_values)) {
  s <- s_values[i]
  chains <- lapply(1:J, function(j) {
    x0 <- runif(1, -5, 5)
    metropolis_hastings(N, s, x0)
  })
  r_hat_values[i] <- calculate_r_hat(chains)
}

rhat_df <- data.frame(s = s_values, R_hat = r_hat_values)
```

```{r, fig.width=8, fig.height=5}
# Plot R-hat values
ggplot(rhat_df, aes(x = s, y = R_hat)) +
  geom_line(color = "purple", size = 1) +
  geom_hline(yintercept = 1.05, color = "red", linetype = "dashed") +
  labs(title = "R-hat Diagnostic Over Proposal Scale (s)",
       x = "Proposal SD (s)", y = "R-hat") +
  theme_minimal()
```

# Conclusion

This R implementation of the Metropolis-Hastings algorithm successfully replicates the statistical behavior and diagnostics of the original Python analysis. The $\hat{R}$ diagnostic highlights the sensitivity of convergence to the choice of proposal distribution scale.
